---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


Analysis of Activity Monitoring Data
====================================

For this assignment we perform some simple analysis of personal
activity monitoring data that records the number of steps for a single
individual in 5-minute intervals over a two month period in October
and November, 2012.

The goal of this assignment is to demonstrate techniques for creating
literate statistical programming documents using *R Markdown* and
*Knitr*. In literate statistical programs explanitory text is
integrated with code chunks in order to tell a complete story of the
data analysis. This is like including very detailed comments in your
analysis scripts, with the additional benefit that they can be
formatted and typeset to create html or other documents.

Knitr is a utility that does this formatting, but also executes the
analysis code at the same time. First, Knitr creates a Markdown
document by running the code and substituting output into the document
in a structured way. Then it can create a html document using
javascript and additionally adding the proper html tags to the
Markdown document.

This report also demonstrates the usage of different plotting methods
to understand and analyse data, as well as a little bit of date object
manipulation. I use the base plotting system in this report, but the
other plotting systems are of course supported.

I use the following library for this analysis, in order to make it
easier to wield
the data: 
```{r}
library(dplyr)
```

## Loading and preprocessing the data 

I first load the data and convert it to a plyr data table. 
The data is downloaded (and unzipped) from the URL: 
[Activity Monitoring Dataset]
(https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip).
The dataset is 52 kB, and contains 17,568 observations of three variables,
* steps: The number of steps taken in a given 5-minute observation.
* date: The date on which the observation was made.
* interval: The 5-minute interval on which the observation was made.

There are 288 unique intervals, ranging from 0 to 2355, so covering a
24-hour period.  There are 61 total days over which the experiment was
performed, but only 53 days that have at least some step data.
Some days have no data at all, for instance, October 1, 2012 and
November 4, 2012.

I coerce the date field to be date objects instead of factors. I
also create a clean dataset that is the original dataset with all of
the missing data removed. Further, we create two grouped datasets,
grouped by date and by interval, for later analysis.


```{r}
dataset <- tbl_df(read.csv('./activity.csv'))
dataset$date <- as.Date(dataset$date)

isMissing <- is.na(dataset$steps)
cleanDataset <- dataset[!isMissing,]
by_date <- group_by(cleanDataset, date)
by_interval <- group_by(cleanDataset, interval)
```

## What is the mean number of steps per day?
First I create a histogram of the total number of steps taken per day. This
is shown in the figure below. Missing data is ignored for this part of the
analysis since there are certain days that have no data.
 
```{r}
spd <- summarize(by_date,spd=sum(steps))
title <- 'Histogram of Number of Steps Per Day'
xlab <- 'Steps/Day'
mean <- mean(spd$spd)
median <- median(spd$spd)
hist(spd$spd, breaks=8, main=title, xlab=xlab)
abline(v=mean, col='blue', lwd=3)
abline(v=median, col='red', lwd=3)
```

The mean number of steps per day is `r mean` and the median number of steps
per day is `r median`. These values are indicated by vertical lines in the
plot above. Since they are very close to eachother, they can not be differentiated
from each other on the plot.

##  What is the average daily activity pattern?
To compute the average pattern of daily activity, namely the number of steps for
each interval averaged over each day, I first group the data by interval using
the clean dataset (with no missing values), and compute the mean of each group.
This is then plotted below.

```{r}
aspi <- summarize(by_interval, aspi=mean(steps))
intervalAtMax <- aspi$interval[which.max(aspi$aspi)]
type='l'
title <- 'Average Number of Steps in Each Interval'
xlab <- 'Interval'
ylab <- 'Average Number of Steps'
plot(aspi$interval, aspi$aspi, type=type, main=title, xlab=xlab, ylab=ylab)
abline(v=intervalAtMax, col='blue', lwd=3)
text(1800, 150, paste('Interval with maximum average \n number of steps: ', 
                      as.character(intervalAtMax)))
```

The interval with the maximum average number of steps is given by `r intervalAtMax`,
corresponding to a maximun value of `r max(aspi$aspi)`.  This interval is indicated 
on the plot with a vertical blue line. 


## Imputing missing values
There are `r numMissingValues` missing observations in this dataset, which will
introduce biases into the analysis. I replace these missing values with the mean
value for that interval, calculated without the missing data (in the previous
section). It is not possible to replace the missing values with the average for
that particular day (another fine strategy) because there are entire days with
missing values.

```{r}
numMissingValues <- sum(isMissing)
for (i in 1:length(dataset$steps)) {
 if (is.na(dataset$steps[i])) {
   dataset$steps[i] <- aspi$aspi[aspi$interval == dataset$interval[i]]
  }
#aspi2 <- summarize(by_interval, aspi2=mean(steps))
#plot(aspi2$interval, aspi2$aspi2, type=type, main=title, xlab=xlab, ylab=ylab)
}
```

## Are there differences in activity patterns between weekdays and weekends?



